---
title: "Biostatistics - Bioinformatics for Computational Genomics"
subtitle: "Final Examination Template"
author: "Chiara Simiele"
date: "Academic Year 2025/2026"
output:
  html_document:
    df_print: paged
    toc: false
    number_sections: false
    theme: cosmo
    highlight: tango
  pdf_document:
    fig_width: 5
    fig_height: 4
geometry: margin=2.5cm
fontsize: 11pt
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

<br>

# Student Information

| Field             | Information |
|:------------------|:------------|
| **Name**          | Chiara      |
| **Surname**       | Simiele     |
| **Student ID**    | 69894A      |
| **Personal Code** | 10769665    |

<br>

------------------------------------------------------------------------

# Instructions

-   Answer **only the two assigned questions**
-   Clearly justify all methodological choices
-   Report relevant statistical results
-   Provide biological and clinical interpretation of findings
-   Include comments directly below each code chunk

<br>

------------------------------------------------------------------------

# Initial Setup

## 1. Load Required Libraries

```{r}
library(survival)
library(survminer)
library(dplyr)
library(stats)
```

## 2. Load the Dataset

```{r}
endometriosis_copy <- read.csv("endometriosis_copy.csv")
endometriosis_gene <- read.csv("endometriosis_gene.csv")
endometriosis_meth <- read.csv("endometriosis_meth.csv")
```

------------------------------------------------------------------------

## 3. Preliminary Data Inspection

```{r}
str(endometriosis_copy)    #312 rows e 24794 columns
str(endometriosis_gene)
str(endometriosis_meth)

head(endometriosis_copy, 1)
head(endometriosis_gene, 1)
head(endometriosis_meth, 1)
```

<br>

------------------------------------------------------------------------

# Question 1

## Assigned Exercise

\*\*Question 1: Investigate the association between disease stage and
diagnosis group, and between diagnosis group and a molecular biomarker:

-   disease_stage (I, II, III, IV) and diagnosis_group (Endometriosis /
    Atypical endometriosis / EAOC).
-   diagnosis_group and DVL1 copy number categorized as copy number loss
    (\<0) and normal/gain (\>=0).

1.  Create contingency tables for both associations.

2.  Test for both associations using chi-squared test and Fisher’s exact
    test. Compare the results and explain when each test is appropriate.
    For the following analyses, recategorize the diagnosis as EAOC
    versus non-EAOC (Endometriosis + Atypical Endometriosis).

3.  Calculate odds ratio (with 95% confidence intervals) comparing
    disease stages using Stage I as the reference category, for the odds
    of belonging to the EAOC group.

4.  Apply Yates’ continuity correction to the 2x2 contingency table
    obtained by comparing advanced disease stages (III-IV) versus early
    disease stages (I, II) with respect to EAOC status (EAOC vs
    non-EAOC).\*\*

------------------------------------------------------------------------

## Step 1 and 2

```{r}

# 1. Create DVL1 categorized variable from copy number data
# Assuming DVL1 column name in endometriosis_copy dataset is cnv_DVL1
endometriosis_copy$dvl1_cat <- factor(
  ifelse(endometriosis_copy$cnv_DVL1 < 0, "Loss (<0)", "Normal/Gain (>=0)")
)


# 2. Contingency Table: disease_stage vs diagnosis_group
table_stage_diag <- table(endometriosis_copy$disease_stage, endometriosis_copy$diagnosis_group)
print(table_stage_diag)

# Tests for Association 1
chisq.test(table_stage_diag)
fisher.test(table_stage_diag, simulate.p.value = TRUE)

# 3. Contingency Table: diagnosis_group vs DVL1 Copy Number
table_dvl1_diag <- table(endometriosis_copy$diagnosis_group, endometriosis_copy$dvl1_cat)
print(table_dvl1_diag)

# Tests for Association 2
chisq.test(table_dvl1_diag)
fisher.test(table_dvl1_diag)
```

### Interpretation

Chi-squared Test vs Fisher's Exact Test: The Pearson's Chi-squared test
evaluates whether an association exists between two categorical
variables by comparing observed counts to expected counts. It relies on
a large-sample asymptotic approximation, requiring (Cochran rule)
expected cell counts to be $\ge 5$ in at least 80% of cells and all
expected frequencies larger than 1. Fisher's Exact Test calculates exact
hypergeometric probabilities without asymptotic assumptions, making it
the preferred method when sample sizes are small or expected counts
don't respect Cochran rules.

Association Results:

-   Association 1: disease_stage vs diagnosis_group. Pearson's
    Chi-squared test yielded $\chi^2 = 99.336$
    ($df = 6, p < 2.2 \times 10^{-16}$). Fisher's Exact Test (with Monte
    Carlo simulation) confirmed a highly significant association
    ($p = 0.0004998$). R issued a warning for the Chi-squared test
    ("Chi-squared approximation may be incorrect") due to several
    zero-count cells in the contingency table which make Chi-squared
    Test unsuitable for a right evaluation of this case. Indeed when
    expected counts in $>20\%$ of cells are $<5$, the asymptotic
    assumption of the Chi-squared test is violated. Therefore, Fisher's
    Exact Test is the appropriate and statistically robust method to
    rely upon. It shows that there is a strong, statistically
    significant association between disease stage and diagnosis group
    ($p < 0.001$). Benign endometriosis and atypical endometriosis cases
    are heavily concentrated in early stages (Stages I and II), whereas
    malignant EAOC cases constitute 100% of the advanced stages (Stages
    III and IV), highlighting a clear pattern of disease severity and
    progression.
-   Association 2: diagnosis_group vs DVL1 Copy Number Status
    (dvl1_cat). Pearson's Chi-squared test resulted in $\chi^2 = 1.4246$
    ($df = 2, p = 0.4905$), and Fisher's Exact Test yielded
    $p = 0.4806$. Both tests yield concordant results and are valid
    because all cells in the $3 \times 2$ contingency table have
    sufficient sample sizes (minimum cell count of 11), well above the
    standard threshold of 5 expected observations per cell. No
    statistically significant association was observed between DVL1 copy
    number loss and diagnostic categories ($p > 0.05$). The proportion
    of DVL1 loss is comparable across all clinical groups (\~48% in
    Atypical Endometriosis, \~42% in EAOC, and \~39% in Endometriosis).
    This suggests that isolated DVL1 copy number loss is not a
    standalone molecular driver of malignant transformation into EAOC in
    this cohort.

------------------------------------------------------------------------

## Step 2

```{r}
# Recategorize diagnosis into binary status: EAOC vs non-EAOC
endometriosis_copy$eaoc_status <- factor(
  ifelse(endometriosis_copy$diagnosis_group == "EAOC", "EAOC", "non-EAOC")
)

# Set "Stage I" as the reference category for disease_stage
endometriosis_copy$disease_stage <- relevel(
  factor(endometriosis_copy$disease_stage), 
  ref = "Stage I"  
)


# Fit binary logistic regression model to estimate odds of belonging to the EAOC group
#To evaluate the association between disease progression and malignant transformation, a binary logistic regression model (glm with family = binomial) was fitted using eaoc_status as the binary response variable ($1 = \text{EAOC}$, $0 = \text{Non-EAOC}$) and disease_stage as the primary categorical/ordinal predictor.

logit_model <- glm(
  eaoc_status ~ disease_stage, 
  data = endometriosis_copy, 
  family = binomial
)


# Extract Odds Ratios (OR) and calculate 95% Confidence Intervals (CI)
or_matrix <- exp(cbind(OR = coef(logit_model), confint(logit_model)))
print(or_matrix)

```

### Interpretation

Binary logistic regression evaluates the odds of having EAOC vs non-EAOC
across disease stages, setting Stage I as the baseline reference
category. Exponentiating the regression coefficients ($\text{e}^\beta$)
yields Odds Ratios (OR) with $95\%$ Confidence Intervals (CI).
Results:The model outputs extreme, unstable estimates for advanced
stages ($\text{OR} \approx 0$, upper bounds reaching Inf or 2.3e+42, and
NA for standard errors). This behavior occurs due to quasi-complete
separation: as shown in the Step 1 contingency table, Stages III and IV
contain exclusively EAOC cases ($100\%$) and zero non-EAOC cases.
Because there are no non-EAOC controls in advanced stages, the maximum
likelihood estimation (MLE) algorithm fails to converge properly for
these parameters, leading to numerical instability and non-calculable
confidence limits. While the uncorrected logistic regression model
breaks down numerically due to empty cells, the underlying biological
pattern is unambiguous: advanced disease stages (III and IV) are
overwhelmingly associated with malignant EAOC status compared to Stage
I.

------------------------------------------------------------------------

## Step 3

```{r}
# Create binary stage variable: Early (Stage I, Stage II) vs Advanced (Stage III, Stage IV)
endometriosis_copy$stage_binary <- factor(
  ifelse(endometriosis_copy$disease_stage %in% c("Stage I", "Stage II"), "Early (I-II)", "Advanced (III-IV)")
) 

# Create 2x2 Contingency Table (Stage Binary vs EAOC Status)
table_2x2 <- table(endometriosis_copy$stage_binary, endometriosis_copy$eaoc_status)
print("--- 2x2 Contingency Table ---")
print(table_2x2)

# Chi-squared test with Yates' continuity correction
chisq_yates <- chisq.test(table_2x2, correct = TRUE)
print("--- Chi-squared Test with Yates' Correction ---")
print(chisq_yates)
```

### Interpretation

*The* $2 \times 2$ contingency table yields 9 EAOC cases and 0 non-EAOC
cases in the Advanced stage (III-IV), compared to 19 EAOC cases and 284
non-EAOC cases in the Early stage (I-II). Pearson's Chi-squared test
with Yates' continuity correction yields $\chi^2 = 82.873$
($df = 1, p < 2.2 \times 10^{-16}$). Yates' Continuity Correction
applied specifically to $2 \times 2$ contingency tables, this correction
subtracts $0.5$ from the absolute difference between observed and
expected frequencies ($\vert{}O - E\vert{} - 0.5$). It adjusts for
approximating a discrete binomial distribution with a continuous
Chi-squared distribution, yielding a more conservative $p$-value to
prevent Type I error (false positives). R displays a warning
("Chi-squared approximation may be incorrect") because one cell in the
$2 \times 2$ table has an observed frequency of $0$ (Advanced non-EAOC),
leading to low expected frequencies in that cell. Despite the
statistical warning, the association between advanced stage disease and
malignant EAOC status remains extremely strong ($p < 0.001$). $100\%$ of
patients presenting with advanced disease (III-IV) belong to the EAOC
malignant group, reinforcing that advanced clinical stages represent a
severe, high-risk phenotypes.

------------------------------------------------------------------------

## Final Comment for Question 1

This multi-step biostatistical analysis provides key insights into risk
stratification for Endometriosis-Associated Ovarian Cancer (EAOC):

-   Clinical Staging and Malignancy: disease stage shows an
    exceptionally strong association with diagnosis status
    ($p < 0.001$). Early stages (I-II) are predominantly composed of
    benign endometriosis and atypical lesions, whereas advanced stages
    (III-IV) consist exclusively of EAOC cases.

-   Methodological Considerations: standard logistic regression models
    encounter quasi-complete separation due to zero control cases in
    advanced stages. Grouping stages into binary categories (early vs
    advanced) and applying Yates' continuity correction
    ($\chi^2 = 82.873, p < 2.2 \times 10^{-16}$) resolves estimation
    issues while confirming a statistically significant risk shift
    toward malignancy in advanced stages.

-   Genomic Biomarkers: isolated copy number loss of the DVL1 gene
    showed no statistically significant association with diagnostic
    groups ($p = 0.4905$). This indicates that DVL1 CNV loss alone is
    insufficient as a standalone biomarker for EAOC transformation and
    must be analyzed alongside broader multi-omic profiles.

<br><br>

------------------------------------------------------------------------

# Question 2

## Assigned Exercise

> Build prognostic models for overall survival in patients affected by
> endometriosis and endometriosis-associated ovarian cancer using Cox
> regression.

1.  Fit a multivariable Cox proportional hazards model including
    clinical predictors such as age at diagnosis, ethnicity, diagnosis
    group (Endometriosis, Atypical endometriosis, EAOC), and relevant
    disease characteristics (bilateral disease). Estimate hazard ratios
    for each prognostic factor.

2.  Test the proportional hazards assumption using Schoenfeld residuals
    or other graphical diagnostics, clearly justifying your choice.

3.  Based on the proportional hazards assumption results, extend the Cox
    model by introducing a stratification according to bilateral disease
    showing evidence of non-proportional / proportional hazards.
    Interpret the results and compare them with the standard Cox model.

4.  Visualize the effect of prognostic factors using adjusted survival
    curves.

------------------------------------------------------------------------

## Step 1

```{r}
# Fit Multivariable Cox Proportional Hazards Model
# Time variable: time_years, Event variable: status (0=alive/censored, 1=death)
cox_model <- coxph( Surv(time_years, status) ~ age + factor(ethnicity) + factor(family_history) + BMI + factor(diagnosis_group) + factor(bilateral_disease), 
  data = endometriosis_copy)

# Summary of Cox Model to extract Hazard Ratios (HR) and 95% Confidence Intervals
summary(cox_model)
```

### Interpretation

The multivariable Cox proportional hazards model demonstrated robust
overall statistical significance across all tests: likelihood Ratio test
($39.32$ on $9\text{ df}, p = 1 \times 10^{-5}$), Wald test ($44.26$ on
$9\text{ df}, p = 1 \times 10^{-6}$), and Score log-rank test ($52.73$
on $9\text{ df}, p = 3 \times 10^{-8}$). Incorporating body mass index
(BMI) improved the model's predictive discrimination, increasing the
Concordance index ($C$-index) to $0.719$ ($\text{SE} = 0.037$).The
estimated Hazard Ratios ($\text{HR} = \exp(\beta)$) for all prognostic
factors are located in the model output under the exp(coef) column, with
their corresponding $95\%$ confidence intervals provided in the lower
section under lower .95 and upper .95.

Evaluation of Prognostic Factors:

-   Diagnosis Group (EAOC): EAOC it's a major negative prognostic
    factor. Patients diagnosed with EAOC exhibited a nearly 3.4-fold
    higher hazard of death compared to the baseline group of atypical
    endometriosis. Benign endometriosis showed no significant survival
    difference relative to baseline ($\text{HR} = 0.95$,
    $95\%\text{ CI } [0.44, 2.05]$, $p = 0.900$).
-   Body Mass Index (BMI): BMI emerged as a statistically significant
    predictor ($\text{HR} = 1.10$, $95\%\text{ CI } [1.03, 1.16]$,
    $p = 0.00218$). Each 1-unit increase in BMI was associated with a
    $9.6\%$ incremental increase in mortality hazard, highlighting the
    adverse prognostic contribution of metabolic factors in this cohort.
-   Other Clinical Covariates: Age at diagnosis
    ($\text{HR} = 1.02, p = 0.177$), family history of cancer
    ($\text{HR} = 0.88, p = 0.681$), presence of bilateral disease
    ($\text{HR} = 0.71, p = 0.255$), and ethnic subgroups ($p > 0.05$
    across all levels relative to Asian baseline) did not show
    statistically significant associations with overall survival.

------------------------------------------------------------------------

## Step 2

```{r}
# Test Proportional Hazards (PH) Assumption using Schoenfeld Residuals
ph_test <- cox.zph(cox_model)
print(ph_test)

# Plot Schoenfeld Residuals for graphical diagnosis
ggcoxzph(ph_test)
```

### Interpretation

The global Schoenfeld test yields a $p$-value of $0.4777$, indicating
that the model as a whole does not exhibit severe global
non-proportionality.

Most predictors satisfy the proportional hazards (PH) assumption: age
($p = 0.8017$), ethnicity ($p = 0.6045$), BMI ($p = 0.6981$),
diagnosis_group ($p = 0.8737$), and bilateral_disease ($p = 0.2466$).

Violation Detected: The covariate family_history displays a
statistically significant violation of the proportional hazards
assumption ($p = 0.0447$). The residual plot reveals a time-dependent
regression coefficient ($\beta(t)$), where the effect of family history
varies over follow-up duration.

Visual inspection of the Schoenfeld residual plots confirms that while
primary clinical predictors (such as diagnosis_group and BMI) show flat
horizontal smoothing curves ($\beta(t) \approx \text{constant}$),
family_history and bilateral_disease display subtle non-linear
trajectories over time. Detecting non-proportional hazards in background
clinical covariates justifies extending the standard multivariable Cox
framework to a stratified Cox model in Step 3. Stratifying by disease
characteristics (bilateral_disease) accounts for underlying baseline
hazard heterogeneity across subgroups, ensuring unbiased and stable
Hazard Ratio estimates for key predictors.

------------------------------------------------------------------------

## Step 3

```{r}
# Fit Stratified Cox Proportional Hazards Model by bilateral_disease
strat_cox_model <- coxph(
  Surv(time_years, status) ~ age + factor(ethnicity) + factor(diagnosis_group) + 
    factor(family_history) + BMI + strata(bilateral_disease), 
  data = endometriosis_copy
)

# Summary of Stratified Cox Model
summary(strat_cox_model)

# Survival curves by stratification variable (bilateral_disease)
fit_strat <- survfit(strat_cox_model)

# Graphical visualization of the two strata baseline curves
ggsurvplot(
  fit_strat,
  data = endometriosis_copy,
  conf.int = FALSE,
  legend.title = "Bilateral Disease",
  legend.labs = c("No (0)", "Yes (1)"),
  xlab = "Time (years)",
  ylab = "Survival Probability",
  ggtheme = theme_minimal(),
  title = "Baseline Survival Probabilities Stratified by Bilateral Disease"
)
# Likelihood Ratio / AIC comparison between models
#Akaike Information Criterion (AIC) is a statistical metric used to compare and select the best model among a set of candidates applied to the same dataset (Lower is Better)
AIC(cox_model, strat_cox_model)
```

### Interpretation

Stratifying by bilateral_disease allows the baseline hazard function
$\lambda_0(t)$ to vary freely across strata while estimating shared
regression coefficients for the remaining covariates. The overall
stratified specification is highly significant ($p = 5 \times 10^{-6}$
on Likelihood Ratio test) with a Concordance index of $C = 0.698$
($\text{SE} = 0.043$).

Prognostic Factor Estimates:

-   Diagnosis Group (EAOC): confirmed as a primary independent risk
    factor. Patients with EAOC exhibit a statistically significant
    $3.24$-fold increase in mortality hazard relative to atypical
    endometriosis ($\text{HR} = 3.24$, $95\%\text{ CI } [1.47, 7.11]$,
    $p = 0.00342$).

-   Body Mass Index (BMI): Maintains independent prognostic significance
    ($\text{HR} = 1.09$, $95\%\text{ CI } [1.03, 1.16]$, $p = 0.00377$),
    indicating a $9.1\%$ relative increase in mortality risk per unit
    increase in BMI.

-   Other Covariates: Age ($\text{HR} = 1.02, p = 0.291$), family
    history ($\text{HR} = 0.95, p = 0.877$), benign endometriosis
    ($\text{HR} = 0.84, p = 0.651$), and ethnicity categories
    ($p > 0.05$ across all levels relative to Asian baseline) show no
    statistically significant associations with overall survival.

Graphical Diagnosis of Stratified Strata: The baseline survival curves
for bilateral disease strata ($0 = \text{Unilateral}$,
$1 = \text{Bilateral}$) exhibit clear temporal divergence, intersecting
around 6–7 years before dropping sharply for the unilateral cohort after
year 9. This crossing of survival trajectories visually confirms
non-proportional hazard behavior between disease extent groups. Allowing
separate stratum-specific baseline hazard functions ($\lambda_{0s}(t)$)
resolves this baseline hazard heterogeneity, directly driving the
substantial model fit improvement ($\Delta\text{AIC} = 76.26$) and
ensuring unbiased hazard ratio estimates for key predictors like EAOC
status and BMI.\*

------------------------------------------------------------------------

## Step 4

```{r}
# Adjusted Survival Curves for BMI

library(survminer)
library(ggplot2)

# Define baseline hypothetical profiles for BMI strata (20, 25, 30)
bmi_df <- data.frame(
  BMI = c(20, 25, 30),
  age = median(endometriosis_copy$age, na.rm = TRUE),
  ethnicity = factor(levels(factor(endometriosis_copy$ethnicity))[1], 
                     levels = levels(factor(endometriosis_copy$ethnicity))),
  diagnosis_group = factor(levels(factor(endometriosis_copy$diagnosis_group))[1], 
                           levels = levels(factor(endometriosis_copy$diagnosis_group))),
  family_history = factor(0, levels = levels(factor(endometriosis_copy$family_history))),
  bilateral_disease = factor(0, levels = levels(factor(endometriosis_copy$bilateral_disease)))
)

# Estimate adjusted survival curves from the multivariable Cox model
fit_bmi <- survfit(cox_model, newdata = bmi_df)

# Render adjusted survival plot for BMI profiles
p_bmi <- ggsurvplot(
  fit_bmi,
  data = endometriosis_copy,
  conf.int = FALSE,
  palette = c("#2ecc71", "#e67e22", "#e74c3c"),
  legend.title = "BMI Profile",
  legend.labs = c("BMI = 20", "BMI = 25", "BMI = 30"),
  xlab = "Time (years)",
  ylab = "Survival Probability",
  ggtheme = theme_minimal(),
  title = "Adjusted Survival Curves by BMI Profiles"
)

p_bmi
```

```{r}

#Adjusted Survival Curves for Diagnosis Group

# Define baseline hypothetical profiles for Diagnosis Group levels
diag_df <- data.frame(
  diagnosis_group = factor(levels(factor(endometriosis_copy$diagnosis_group)), 
                           levels = levels(factor(endometriosis_copy$diagnosis_group))),
  age = median(endometriosis_copy$age, na.rm = TRUE),
  ethnicity = factor(levels(factor(endometriosis_copy$ethnicity))[1], 
                     levels = levels(factor(endometriosis_copy$ethnicity))),
  family_history = factor(0, levels = levels(factor(endometriosis_copy$family_history))),
  bilateral_disease = factor(0, levels = levels(factor(endometriosis_copy$bilateral_disease))),
  BMI = median(endometriosis_copy$BMI, na.rm = TRUE)
)

# Estimate adjusted survival curves from the multivariable Cox model
fit_diag <- survfit(cox_model, newdata = diag_df)

# Render adjusted survival plot for Diagnosis Groups
p_diag <- ggsurvplot(
  fit_diag,
  data = endometriosis_copy,
  conf.int = FALSE,
  palette = c("#3498db", "#9b59b6", "#c0392b"),
  legend.title = "Diagnosis",
  legend.labs = levels(factor(endometriosis_copy$diagnosis_group)),
  xlab = "Time (years)",
  ylab = "Survival Probability",
  ggtheme = theme_minimal(),
  title = "Adjusted Survival Curves by Diagnosis Group"
)

p_diag
```

### Interpretation

To evaluate the isolated prognostic impact of primary risk factors while
controlling for baseline confounding, adjusted survival probability
curves were generated for target clinical profiles using the fitted
multivariable Cox model (survfit with newdata). Confounding background
covariates were fixed at their sample medians (for continuous variables)
or reference baseline categories (for factor variables).

-   Body Mass Index (BMI) Risk Gradient: the adjusted survival curves
    across distinct BMI strata ($\text{BMI} = 20$, $25$, and $30$)
    demonstrate a clear, monotonic dose-response relationship over time.
    Higher baseline BMI profiles consistently shift the survival curve
    downward: at 5 years, survival probability drops from approximately
    $92\%$ for normal weight ($\text{BMI} = 20$) to nearly $75\%$ for
    obese profiles ($\text{BMI} = 30$), reaching an even wider gap at 10
    years ($\sim 60\%$ vs $\sim 30\%$). This visually confirms the
    significant incremental hazard ($\text{HR} = 1.096$ per BMI unit,
    $p = 0.00218$) identified in the multivariable model.

-   Diagnosis group: the adjusted survival curves illustrate a
    pronounced survival disadvantage for patients diagnosed with EAOC
    (purple curve). While benign endometriosis (red curve) and atypical
    endometriosis (blue curve) exhibit virtually overlapping, favorable
    survival trajectories ($\sim 85\%$ survival at 5 years and
    $\sim 50\%$ at 10 years), the EAOC trajectory declines sharply,
    dropping to $\sim 60\%$ at 5 years and falling below $15\%$ at 10
    years. This visual separation aligns with the $3.39$-fold mortality
    hazard increase ($\text{HR} = 3.39, p = 0.00205$) observed for EAOC,
    while confirming no distinct risk difference between benign and
    atypical endometriosis ($\text{HR} = 0.95, p = 0.900$).

## Final Comment for Question 2

This multivariable survival analysis evaluated the prognostic drivers of
overall mortality in patients with endometriosis and associated ovarian
cancers (EAOC), while addressing model assumptions through advanced
stratification techniques. Multivariable Cox regression identified EAOC
diagnosis and Body Mass Index (BMI) as the primary independent
predictors of overall mortality. Age, family history, and ethnic
subgroups demonstrated no statistically significant association with
overall survival.

Initial proportional hazard diagnostics revealed temporal
non-proportionality. Extending the model to a stratified Cox framework
by bilateral_disease allowed stratum-specific baseline hazard functions
$\lambda_{0s}(t)$ while preserving common hazard ratios for remaining
covariates. Although this adaptation slightly reduced the apparent
concordance index ($C = 0.698$ vs $0.721$), it yielded a massive
goodness-of-fit improvement, reducing the Akaike Information Criterion.
Adjusted survival probability profiles (survfit with target newdata)
visually isolated the long-term impact of key predictors. The adjusted
curves demonstrate a severe, continuous survival drop for the EAOC
cohort and a clear dose-response risk gradient across BMI strata.

In conclusion, the stratified Cox proportional hazards model provides a
robust, methodologically sound framework for survival prediction in this
population. These findings underscore the critical need for early
oncological monitoring upon malignant transformation to EAOC and
highlight metabolic control (BMI management) as a key targeted strategy
in long-term patient care.

<br><br>

------------------------------------------------------------------------
